Causal graph analysis of COVID-19 observational data in German districts reveals effects of mobility, awareness, and temperature

### 2021-03-01 manuscript/v2_medrxiv/manuscript.Rmd
small change in title to test difflatex

---
title: Causal analysis of COVID-19 observational data in German districts reveals effects of mobility, awareness, and temperature
author:
  - name: Edgar Steiger
    email: esteiger@zi.de
    affiliation: Zi
  - name: Tobias Mußgnug
    email: tmussgnug@zi.de
    affiliation: Zi
  - name: Lars Eric Kroll
    email: lkroll@zi.de
    affiliation: Zi
    corresponding: lkroll@zi.de
address:
  - code: Zi
    address: Central Research Institute of Ambulatory Health Care in Germany (Zi), Salzufer 8, D-10587 Berlin, Germany
abstract: Mobility, awareness, and weather are suspected to be causal drivers for new cases of COVID-19 infection. Correcting for possible confounders, we estimated their causal effects on reported case numbers. To this end, we used a directed acyclic graph (DAG) as a graphical representation of the hypothesized causal effects of the aforementioned determinants on new reported cases of COVID-19. Based on this, we computed valid adjustment sets of the possible confounding factors. We collected data for Germany from publicly available sources (e.g. Robert Koch Institute, Germany's National Meteorological Service, Google) for 401 German districts over the period of 15 February to 8 July 2020, and estimated total causal effects based on our DAG analysis by negative binomial regression. Our analysis revealed favorable causal effects of increasing temperature, increased public mobility for essential shopping (grocery and pharmacy), and awareness measured by COVID-19 burden, all of them reducing the outcome of newly reported COVID-19 cases. Conversely, we saw adverse effects of public mobility in retail and recreational areas, awareness measured by searches for “corona” in Google, and higher rainfall, leading to an increase in new COVID-19 cases. This comprehensive causal analysis of a variety of determinants affecting COVID-19 progression gives strong evidence for the driving forces of mobility, public awareness, and temperature, whose implications need to be taken into account for future decisions regarding pandemic management.

>>>>

---
title: Causal graph analysis of COVID-19 observational data in German districts reveals effects of mobility, awareness, and temperature
author:
  - name: Edgar Steiger
    email: esteiger@zi.de
    affiliation: Zi
  - name: Tobias Mußgnug
    email: tmussgnug@zi.de
    affiliation: Zi
  - name: Lars Eric Kroll
    email: lkroll@zi.de
    affiliation: Zi
    corresponding: lkroll@zi.de
address:
  - code: Zi
    address: Central Research Institute of Ambulatory Health Care in Germany (Zi), Salzufer 8, D-10587 Berlin, Germany
abstract: Mobility, awareness, and weather are suspected to be causal drivers for new cases of COVID-19 infection. Correcting for possible confounders, we estimated their causal effects on reported case numbers. To this end, we used a directed acyclic graph (DAG) as a graphical representation of the hypothesized causal effects of the aforementioned determinants on new reported cases of COVID-19. Based on this, we computed valid adjustment sets of the possible confounding factors. We collected data for Germany from publicly available sources (e.g. Robert Koch Institute, Germany's National Meteorological Service, Google) for 401 German districts over the period of 15 February to 8 July 2020, and estimated total causal effects based on our DAG analysis by negative binomial regression. Our analysis revealed favorable causal effects of increasing temperature, increased public mobility for essential shopping (grocery and pharmacy), and awareness measured by COVID-19 burden, all of them reducing the outcome of newly reported COVID-19 cases. Conversely, we saw adverse effects of public mobility in retail and recreational areas, awareness measured by searches for “corona” in Google, and higher rainfall, leading to an increase in new COVID-19 cases. This comprehensive causal analysis of a variety of determinants affecting COVID-19 progression gives strong evidence for the driving forces of mobility, public awareness, and temperature, whose implications need to be taken into account for future decisions regarding pandemic management.

### 2021-03-01 manuscript/v2_medrxiv/manuscript.tex
small change in title to test difflatex

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{natbib}
\bibliographystyle{plainnat}
\usepackage{longtable}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Causal analysis of COVID-19 observational data in German districts reveals effects of mobility, awareness, and temperature},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

@@

  \title{Causal analysis of COVID-19 observational data in German districts reveals effects of mobility, awareness, and temperature}
    \author[Zi]{Edgar Steiger}
   \ead{esteiger@zi.de} 
    \author[Zi]{Tobias Mußgnug}
   \ead{tmussgnug@zi.de} 
    \author[Zi]{Lars Eric Kroll}
   \ead{lkroll@zi.de} 
      \address[Zi]{Central Research Institute of Ambulatory Health Care in Germany (Zi), Salzufer 8, D-10587 Berlin, Germany}

@@

Within our framework, we saw significant effects for mobility in retail/recreational areas and essential shopping (grocery and pharmacy). In the former, an increase of 1 percent point mobility compared to the reference period (03 January to 06 February 2020) leads to an increase of the daily reported case number by about 0.8 percent. Contrarily, a corresponding increase of 1 percent point for the areas of grocery/pharmacy leads to a decrease in the reported case number by approximately 0.5 percent. Mobility on workplaces showed a small effect of 0.3 increase in case numbers for every 1 percent point increase in mobility. Other causal effects of mobility were insubstantial and not consistent in their direction (99 percent confidence intervals of estimates include 1). Figure \ref{fig:effectsrange} shows the effects of mobility on a range of possible values. Thus, we expect an increase of daily cases by approximately 23 percent if mobility in retail/recreation reaches baseline levels of 0 percent difference to the reference period. On the other hand, an increase of mobility for grocery/pharmacy by 10 percent points compared to the reference period leads to a reduction of the infection rate by approximately 7 percent.

``Awareness'' had two opposite effects on the outcome in our DAG. Awareness measured by Google searches for \emph{corona} had a positive effect on the number of reported cases. An one percent point increase of the state's Google searches (relative to other states and the observation period) leads to an increase of approximately 1 percent. For example, if a district shows 10 percent points more relative searches for \emph{corona} then another one, we expect approximately 11 percent more infections for this district after 8 days. \emph{COVID-19 burden} (reported number of cases on day of exposure) affected the outcome negatively, where every additional daily case in the district leads to a 0.6 percent decrease in newly reported case numbers. The last plot in Figure \ref{fig:effectsrange} visualizes this relationship: For a local outbreak with 20 daily cases as COVID-19 burden, we estimate as total causal effect a subsequent reduction of infection rate by 9 percent.

Within our model, we observed a causal effect of temperature and all other weather variables. Every increase of 1 degree Celsius in temperature leads to a reduction of the daily reported case numbers by approximately 0.8 percent. On the other hand, we found an increasing effect of rainfall: One millimeter (=1 liter per square meter) more rainfall leads to an increase of reported case numbers by approximately 1.1 percent. We observe small effects for humidity and wind as well (higher humidity leading to fewer cases, stronger wind leading to less cases). In perspective (Figure \ref{fig:effectsrange}), with temperature we expect an increase by approximately 9 percent at a daily average temperature of \(0^{\circ}C\). For rainfall, we expect on a very rainy day with 10 mm rainfall a corresponding increase of the infection rate by approximately 2 percent.

In all cases we opted to use the reduced optimal adjustment set over the minimal adjustment sets because of higher pseudo-\(R^2\) values (mostly above 0.3), except for mobility, where the minimal adjustment set had a higher pseudo-\(R^2\). Notably, these sets always include most of our socio-demographic variables as confounders (cf.~Table \ref{tab:adjsets}) as well as the policy variables (except for COVID-19 burden). We also decided for the lag of 8 days based on the highest pseudo-\(R^2\) values compared to other lags on the chosen adjustment sets. Similarly, negative binomial regression was chosen over Poisson regression, because the latter showed overdispersion and overall lower pseudo-\(R^2\) values.

@@

In addition, in alignment with other recent published studies, our results confirm evidence which associated a negative effect of temperature on new COVID-19 cases \citep{bannister-tyrrell_preliminary_2020, demongeot_temperature_2020, liu_impact_2020, qi_covid-19_2020, shi_impact_2020, sobral_association_2020, tosepu_correlation_2020, Wang2020temperature, wu_effects_2020}. It is however controversial to other scientific literature describing no effects \citep{briz-redon_spatio-temporal_2020, iqbal_nexus_2020, jahangiri_sensitivity_2020, juni_impact_2020, yao_no_2020} or even converse correlations \citep{auler_evidence_2020, xie_association_2020}. The conflicting results might be explained by different climates and characteristics of the populations under study. While we are confident that our strict causal analysis resulted in effect estimates as undistorted as possible, there might be unconsidered bias in those other studies. Further research needs to be done to elucidate the biological characteristics of the novel virus SARS-CoV-2 regarding its ambient temperature survival and transmission. Finally, we found a positive causal effect of increment precipitation and a raise in COVID-19 cases, which supports previous observations \citep{sobral_association_2020}.

@@

While use of a causal DAG is itself a strong tool to identify causal effects (and not just statistical associations), it introduces two limitations: causal assumptions within the graph (depicted by edges) need to be well justified, and the statistical regression model that calculates total causal effects needs to be appropriate for the task at hand. We endorse our graph as a basis for discussion on residual confounding. We did not try to construct the DAG from the available data (cf.~\citep{gencoglu2020causal}). As such, our proposed DAG is not entirely consistent with the data and there are conditional dependencies between variables that cannot be dissolved by adding edges to the DAG (e.g.~between the policies like contact restrictions and mandatory face masks). Another way to identify potential problems in the proposed DAG is to perform a sensitivity analysis of its structure by inspecting its maximal ancestral graph (MAG) or its Markov equivalence class represented by a complete partially DAG (CPDAG) and the existence of valid adjustment sets for these generalized graphs \citep{perkovic2017complete}. For the MAG derived from our DAG, only the effects for exposures mobility and searches for corona can be estimated with valid adjustment sets, while for the Markov equivalence class all exposures but COVID-19 burden lead to valid adjustments sets. A further analysis of these implications is out of the scope of this paper.

We observed overdispersion and a substantial increase in model performance with a negative binomial regression compared to Poisson regression, which is in line with the results on COVID-19 daily case counts of \citet{kraemer_effect_2020} and others \citep{liu_impact_2020, bannister-tyrrell_preliminary_2020, Islamm2743}. We did not model case counts with a differential equation model like the classic SIR-model \citep{kermack1991contributions} and its successors, since these are more suited to prediction \citep[e.g.~][]{rkimodellierung} while our choice of a negative binomial regression framework allowed us to estimate the effects of confounders more reliably. There are more advanced statistical methods for count data, e.g.~zero-inflated models and mixed models. We tested both approaches as extensions to the negative binomial regression and experienced numerical problems and increased computing time, along with an insubstantial increase in model performance. Furthermore, our model assumed that all variables have effects proportional to the size of their measurements. It is possible that some variables show saturation effects or opposite effects for lower, medium, or high values. This could be modelled with polynomial or other transformations of the variables, which we did not employ due to limited temporal and spatial data availability. Use of a fixed DAG with effect estimation via regression assumes that data was generated by the same underlying process for the observation period. By inclusion of the successive mitigation policies as binary variables we were able to explain some of the variance caused by the changing dynamics of case numbers (similar to \citep{Islamm2743}).

@@

The causal effects of awareness measured via searches for ``corona'' and the COVID-19 burden are harder to interpret. We assume that within our model, the searches for ``corona'' are an insufficient proxy for awareness, while the decreasing effect for future case numbers of high daily COVID-19 burden numbers indicates it affects individual risk-behavior as a deterrent.

Similarly, the effects of temperature and rainfall can be interpreted as causal effects for indoor and outdoor activities, such that higher temperatures and low rainfall indicate more people spending time outdoor while lower temperatures and high rainfall result in indoor activities, which lead to more infections. In this light, we advocate for precautious measures like increased hygiene, face masks, and air ventilation for unavoidable indoor activities.

@@

Our findings suggest that the causal effects of mobility, awareness, and weather need to be taken strongly into account when deciding for mitigation and suppression measures, depending on the recent and future COVID-19 pandemic development.

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{natbib}
\bibliographystyle{plainnat}
\usepackage{longtable}
\usepackage{graphicx}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Causal graph analysis of COVID-19 observational data in German districts reveals effects of mobility, awareness, and temperature},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

% Pandoc citation processing

@@

  \title{Causal graph analysis of COVID-19 observational data in German districts reveals effects of mobility, awareness, and temperature}
    \author[Zi]{Edgar Steiger}
   \ead{esteiger@zi.de} 
    \author[Zi]{Tobias Mußgnug}
   \ead{tmussgnug@zi.de} 
    \author[Zi]{Lars Eric Kroll}
   \ead{lkroll@zi.de} 
      \address[Zi]{Central Research Institute of Ambulatory Health Care in Germany (Zi), Salzufer 8, D-10587 Berlin, Germany}

@@

Within our framework, we saw significant effects for mobility in retail/recreational areas and essential shopping (grocery and pharmacy). For retail/recreation, an increase of 1 percent point mobility compared to the reference period (03 January to 06 February 2020) leads to an increase of the daily reported case number by about 0.8 percent. Contrarily, a corresponding increase of 1 percent point for the areas of grocery/pharmacy leads to a decrease in the reported case number by approximately 0.5 percent. Mobility on workplaces showed a small effect of 0.3 increase in case numbers for every 1 percent point increase in mobility. Other causal effects of mobility were insubstantial and not consistent in their direction (99 percent confidence intervals of estimates include 1). Figure \ref{fig:effectsrange} shows the effects of mobility on a range of possible values. Thus, we expect an increase of daily cases by approximately 23 percent if mobility in retail/recreation reaches baseline levels of 0 percent difference to the reference period. On the other hand, an increase of mobility for grocery/pharmacy by 10 percent points compared to the reference period leads to a reduction of the infection rate by approximately 7 percent.

``Awareness'' had two opposite effects on the outcome in our DAG. Awareness measured by Google searches for \emph{corona} had a positive effect on the number of reported cases. An one percent point increase of the state's Google searches (relative to other states and the observation period) leads to an increase of approximately 1 percent. For example, if a district shows 10 percent points more relative searches for \emph{corona} than another one, we expect approximately 11 percent more infections for this district after 8 days. \emph{COVID-19 burden} (reported number of cases on day of exposure) affected the outcome negatively, where every additional daily case in the district leads to a 0.6 percent decrease in newly reported case numbers. The last plot in Figure \ref{fig:effectsrange} visualizes this relationship: For a local outbreak with 20 daily cases as COVID-19 burden, we estimate as total causal effect a subsequent reduction of infection rate by 11 percent.

Within our model, we observed a causal effect of temperature and all other weather variables. Every increase of 1 degree Celsius in temperature leads to a reduction of the daily reported case numbers by approximately 0.8 percent. On the other hand, we found an increasing effect of rainfall: One millimeter (=1 liter per square meter) more rainfall leads to an increase of reported case numbers by approximately 1.1 percent. We observe effects for humidity and wind as well (higher humidity leading to more cases, stronger wind leading to less cases). In perspective (Figure \ref{fig:effectsrange}), with temperature we expect an increase by approximately 17.8 percent at a daily average temperature of \(0^{\circ}C\) compared to a day with \(20^{\circ}C\). For rainfall, we expect on a rainy day with 10 mm rainfall a corresponding increase of the infection rate by approximately 11.4 percent compared to a day with no precipitation.

In all cases we opted to use the reduced optimal adjustment set over the minimal adjustment sets because of higher pseudo-\(R^2\) values (mostly above 0.3), except for mobility, where the minimal adjustment set had a higher pseudo-\(R^2\). Notably, these sets always include most of our socio-demographic variables as confounders as well as the policy variables (cf.~Table \ref{tab:adjsets}, with the exception being COVID-19 burden). We also decided for the lag of 8 days based on the highest pseudo-\(R^2\) values compared to other lags on the chosen adjustment sets. Similarly, negative binomial regression was chosen over Poisson regression, because the latter showed overdispersion and overall lower pseudo-\(R^2\) values.

@@

In addition, in alignment with other recent published studies, our results confirm evidence which associated a negative effect of temperature on new COVID-19 cases \citep{bannister-tyrrell_preliminary_2020, demongeot_temperature_2020, liu_impact_2020, qi_covid-19_2020, shi_impact_2020, sobral_association_2020, tosepu_correlation_2020, Wang2020temperature, wu_effects_2020}. It is however controversial to other scientific literature describing no effects \citep{briz-redon_spatio-temporal_2020, iqbal_nexus_2020, jahangiri_sensitivity_2020, juni_impact_2020, yao_no_2020} or even converse correlations \citep{auler_evidence_2020, xie_association_2020}. The conflicting results might be explained by different climates and characteristics of the populations under study. While we are confident that our strict causal analysis resulted in effect estimates as undistorted as possible, there might be unconsidered bias in those other studies. Further research needs to be done to elucidate the biological characteristics of the novel virus SARS-CoV-2 regarding its ambient temperature survival and transmission. Finally, we found a positive causal effect of increased precipitation and a raise in COVID-19 cases, which supports previous observations \citep{sobral_association_2020}.

@@

While use of a causal DAG is itself a strong tool to identify causal effects (and not just statistical associations), it introduces two limitations: causal assumptions within the graph (depicted by edges) need to be well justified, and the statistical regression model that calculates total causal effects needs to be appropriate for the task at hand. We endorse our graph as a basis for discussion on residual confounding. We did not try to construct the DAG from the available data (cf. \citep{gencoglu2020causal}). As such, our proposed DAG is not entirely consistent with the data and there are conditional dependencies between variables that cannot be dissolved by adding edges to the DAG (e.g.~between the policies like contact restrictions and mandatory face masks). Another way to identify potential problems in the proposed DAG is to perform a sensitivity analysis of its structure by inspecting its maximal ancestral graph (MAG) or its Markov equivalence class represented by a complete partially DAG (CPDAG) and the existence of valid adjustment sets for these generalized graphs \citep{perkovic2017complete}. For the MAG derived from our DAG, only the effects for exposures mobility and searches for corona can be estimated with valid adjustment sets, while for the Markov equivalence class all exposures but COVID-19 burden lead to valid adjustments sets. A further analysis of these implications is out of the scope of this paper.

We observed overdispersion and a substantial increase in model performance with a negative binomial regression compared to Poisson regression, which is in line with the results on COVID-19 daily case counts of \citet{kraemer_effect_2020} and others \citep{liu_impact_2020, bannister-tyrrell_preliminary_2020, Islamm2743}. We did not model case counts with a differential equation model like the classic SIR-model \citep{kermack1991contributions} and its successors, since these are more suited to prediction \citep[e.g.][]{rkimodellierung} while our choice of a negative binomial regression framework allowed us to estimate the effects of confounders more reliably. There are more advanced statistical methods for count data, e.g.~zero-inflated models and mixed models. We tested both approaches as extensions to the negative binomial regression and experienced numerical problems and increased computing time, along with an insubstantial increase in model performance. Furthermore, our model assumed that all variables have effects proportional to the size of their measurements. It is possible that some variables show saturation effects or opposite effects for low, medium, or high values. This could be modelled with polynomial or other transformations of the variables, which we did not employ due to limited temporal and spatial data availability. Use of a fixed DAG with effect estimation via regression assumes that data was generated by the same underlying process for the observation period. By inclusion of the successive mitigation policies as binary variables we were able to explain some of the variance caused by the changing dynamics of case numbers (similar to \citep{Islamm2743}).

@@

The causal effects of awareness measured via searches for ``corona'' and the COVID-19 burden are harder to interpret. We assume that within our model, the searches for ``corona'' are an insufficient proxy for awareness, while the decreasing effect for future case numbers of high daily COVID-19 burden indicates it affects individual risk-behavior and entails effective non-pharmaceutical interventions.

Similarly, the effects of temperature and rainfall can be interpreted as causal effects for indoor and outdoor activities, such that higher temperatures and low rainfall indicate more people spending time outdoor while lower temperatures and high rainfall result in indoor activities, which lead to more infections. Current research suggests this to be due to the prevalent airborne and respiratory droplets and aerosol transmission of the SARS-CoV-2 virus \citep{world2020transmission}. In this light, we advocate for precautious measures like increased hygiene, face masks, and air ventilation for unavoidable indoor activities.

@@

Our findings suggest that the causal effects of mobility, awareness, and weather need to be taken into account when deciding for mitigation and suppression measures, depending on the recent and future COVID-19 pandemic development.
